Circumnuclear Media and Accretion Rates of Quiescent Supermassive Black Holes

### 2015-04-07 ms.tex
fixed Ia radius

To within a constant of order unity, equation (\ref{eq:cooling2}) also
equals the ratio of the gas cooling timescale $t_{\rm cool} \equiv (3n
kT/2 \mu)/\dot{q}_{\rm rad}$ to the free-fall time $t_{\rm ff}$ at the
stagnation radius, as can be derived using the equality

\begin{align}
\rho|_{r_{\rm s}}=\frac{3 q\tff|_{r_{\rm s}}}{2-\Gamma}.
\label{eq:rhors2}
\end{align}
that results by combining equations
\eqref{eq:mdot_analytic},\eqref{eq:mdot_gas}, and
\eqref{eq:rhostarrs}.  \citet{McCourt+12} argue cooling instability
develops in a hydrostatic atmosphere if $t_{\rm cool} \lesssim 10 t_{\rm ff}$,
so equation (\ref{eq:cooling2}) represents a good proxy for
instability in this case as well (see also \citealt{Li&Bryan14a}, \citealt{Li&Bryan14b}).

@@

However, even if SN Ia are rare near the stagnation radius, SN Ia may cap the
SMBH accretion rate by periodically blowing gas out of the nucleus of
low mass galaxies.  Between successive SNe, stars release a
gaseous mass $M_{\rm g} \approx \eta M_{\star}\tau_{\rm Ia}/t_{\rm h}$
interior to the Ia radius, which is locally gravitationally bound to the SMBH
by an energy $E_{\rm bind} \sim M_{\rm g}\sigma_0^{2}\mathcal{F}^{2}$, where typically $\mathcal{F} 
\lesssim$ few.  From the above definitions we find that 
\be
\frac{E_{\rm Ia}}{E_{\rm bind}} \sim \frac{(v_{w}^{\rm Ia})^{2}}{2\mathcal{F}^{2}\sigma_0^{2}}.  
\label{eq:blowout}
\ee
Hence, for low mass BHs with $\sigma_0 \ll \mathcal{F}v^{\rm Ia}_{w}/2$, SN Ia are capable of dynamically clearing out gas from radii $\sim r_{\rm Ia} \gtrsim r_{\rm s}$.  Even when heating is sufficiently weak that the stagnation radius formally exceeds $r_{\rm Ia}$, the SMBH accretion rate is still potentially limited to a value
\begin{eqnarray}
\frac{\dot{M}_{\rm Ia}}{\dot{M}_{\rm edd}} &\approx& \frac{\eta M_{\bullet}}{\dot{M}_{\rm edd} t_{\rm h}}\left(\frac{r_{\rm Ia}}{r_{\rm inf}}\right)^{2-\Gamma} \approx \nonumber \\
 && \begin{cases}
    4.5 \times 10^{-4} M_{\bullet,8}^{-1.33}(\tau_{\star}/t_{\rm h})^{-0.2}
   & \text{core} \\
    2.2 \times 10^{-4} \Mbheight^{-0.84}(\tau_{\star}/t_{\rm h})^{-0.6}   & \text{cusp},
  \end{cases}
  \label{eq:eddr_Ia}
\end{eqnarray}
obtained substituting the Ia radius $r_{\rm Ia}$ (eq.~[\ref{eq:rIa}])
for $r_{\rm s}$ in the derivation leading to our estimate of $\dot{M}$
(eq.~[$\ref{eq:eddr_analytic}$]).  Note, however, that for low mass
SMBHs, $\dot{M}_{\rm Ia}$ generally exceeds the maximum accretion rate
for a thermally stable flow $\dot{M}_{\rm TI}$
(eq.~[\ref{eq:Mdotmax}]), implying that dynamical `blow-out' from SNe
Ia cannot alone prevent cooling instabilities in the nuclei of low
mass galaxies {\bf AG--careful one needs a separate criterion for
  thermal stability when heating is dominated by the Ias}.

The deep gravitational potential wells of high mass galaxies $\sigma_0
\gg f v^{\rm Ia}_{w}$ prevent SN Ia from dynamically clearing out gas
in these systems.  Equation (\ref{eq:eddr_Ia}) nevertheless represents
a cap on the accretion rate in practice because if other source of
heating are ineffective, then the Ia heating rate (eq.~[\ref{eq:vIa}])
is usually sufficiently high to prevent the stagnation radius
(calculated including the heating from SN Ia) from substantially
exceeding $r_{\rm Ia}$.

If Ia supernovae are the dominant source of heating we expect $\rs\sim
r_{\rm Ia}$. If the stagnation radius moves inwards from $\rIa$, the
decreased heating will force it outwards again. If the it moves
outwards the Ia heating will force it inwards. Thus,
equations~\eqref{eq:cooling2} and ~\eqref{eq:Mdotmax} will no longer
be valid as they assume that $\rs$ follows equation
\eqref{eq:stag_simple}. We derive $\dot{q}_{\rm
  heat}/\left|\dot{q}_{\rm rad}\right|$ for $\rs=\rIa$ to evaluate
thermal stability in SNe Ia dominate the heating.

\begin{align}
&\rho(\rIa)\simeq \frac{3 q(\rIa)}{2-\Gamma} \nonumber\\
&T(\rIa)\simeq \frac{\gamma-1}{\gamma}\frac{\mu m_p v^{\rm Ia \,\,
    2}_{w}}{2k}\nonumber\\
&\left.\frac{\dot{q}_{\rm heat}}{\left|\dot{q}_{\rm rad}\right|}\right|_{\rIa}\simeq
\begin{cases}
  90 \Mbheight^{1.3} (\epsilon_{\rm Ia}/0.4)^{1.7} (\tau_{\star}/\th)^{-3}
  \eta_{0.02}^{-2.7} & \text{core}\nonumber\\
   90 \Mbheight^{0.8} (\epsilon_{\rm Ia}/0.4)^{1.7} (\tau_{\star}/\th)^{-2.7}
  \eta_{0.02}^{-2.7} & \text{cusp}\nonumber\\
\end{cases}\\&\simeq
\begin{cases}
  90 \Mbheight^{1.3} (\epsilon_{\rm Ia}/0.4)^{1.7}
  (\tau_{\star}/\th)^{0.5}
 & \text{core}\\
   90 \Mbheight^{0.8} (\epsilon_{\rm Ia}/0.4)^{1.7} (\tau_{\star}/\th)^{0.9}
  & \text{cusp}\\
\end{cases}
\end{align}

>>>>

To within a constant of order unity, equation (\ref{eq:cooling2}) also
equals the ratio of the gas cooling timescale $t_{\rm cool} \equiv (3n
kT/2 \mu)/\dot{q}_{\rm rad}$ to the free-fall time $t_{\rm ff}$ at the
stagnation radius, as can be derived using the equality
\begin{align}
\rho|_{r_{\rm s}}=\frac{3 q\tff|_{r_{\rm s}}}{2-\Gamma}.
\label{eq:rhors2}
\end{align}
that results by combining equations
\eqref{eq:mdot_analytic},\eqref{eq:mdot_gas}, and
\eqref{eq:rhostarrs}.  \citet{McCourt+12} argue cooling instability
develops in a hydrostatic atmosphere if $t_{\rm cool} \lesssim 10 t_{\rm ff}$,
so equation (\ref{eq:cooling2}) represents a good proxy for
instability in this case as well (see also \citealt{Li&Bryan14a}, \citealt{Li&Bryan14b}).

@@

However, even if SN Ia are rare near the stagnation radius, SN Ia may cap the
SMBH accretion rate by periodically blowing gas out of the nucleus of
low mass galaxies.  Between successive SNe, stars release a
gaseous mass $M_{\rm g} \approx \eta M_{\star}\tau_{\rm Ia}/t_{\rm h}$
interior to the Ia radius, which is locally gravitationally bound to the SMBH
by an energy $E_{\rm bind} \sim M_{\rm g}\sigma_0^{2}\mathcal{F}^{2}$, where typically $\mathcal{F} 
\lesssim$ few.  From the above definitions we find that 
\be
\frac{E_{\rm Ia}}{E_{\rm bind}} \sim \frac{(v_{w}^{\rm Ia})^{2}}{2\mathcal{F}^{2}\sigma_0^{2}}.  
\label{eq:blowout}
\ee
Hence, for low mass BHs with $\sigma_0 \ll v^{\rm Ia}_{w}$, SN Ia are capable of dynamically clearing out gas from radii $\sim r_{\rm Ia} \gtrsim r_{\rm s}$.  Even when heating is sufficiently weak that the stagnation radius formally exceeds $r_{\rm Ia}$, the SMBH accretion rate is still potentially limited to a value
\begin{eqnarray}
\frac{\dot{M}_{\rm Ia}}{\dot{M}_{\rm edd}} &\approx& \frac{\eta M_{\bullet}}{\dot{M}_{\rm edd} t_{\rm h}}\left(\frac{r_{\rm Ia}}{r_{\rm inf}}\right)^{2-\Gamma} \approx \nonumber \\
 && \begin{cases}
    4.5 \times 10^{-4} M_{\bullet,8}^{-1.33}(\tau_{\star}/t_{\rm h})^{-0.2}
   & \text{core} \\
    2.2 \times 10^{-4} \Mbheight^{-0.84}(\tau_{\star}/t_{\rm h})^{-0.6}   & \text{cusp},
  \end{cases}
  \label{eq:eddr_Ia}
\end{eqnarray}
obtained substituting the Ia radius $r_{\rm Ia}$ (eq.~[\ref{eq:rIa}])
for $r_{\rm s}$ in the derivation leading to our estimate of $\dot{M}$
(eq.~[$\ref{eq:eddr_analytic}$]).  
%Note, however, that for low mass
%SMBHs, $\dot{M}_{\rm Ia}$ generally exceeds the maximum accretion rate
%for a thermally stable flow $\dot{M}_{\rm TI}$
%(eq.~[\ref{eq:Mdotmax}]), implying that dynamical `blow-out' from SNe
%Ia cannot alone prevent cooling instabilities in the nuclei of low
%mass galaxies.

The deep gravitational potential wells of high mass galaxies $\sigma_0
\gg  v^{\rm Ia}_{w}$ prevent SN Ia from dynamically clearing out gas
in these systems.  Equation (\ref{eq:eddr_Ia}) nevertheless represents
a cap on the accretion rate in practice because if other source of
heating are ineffective, then the Ia heating rate (eq.~[\ref{eq:vIa}])
is usually sufficiently high to prevent the stagnation radius
(calculated including the heating from SN Ia) from substantially
exceeding $r_{\rm Ia}$. Stated another way, if the stagnation radius moves inwards from $\rIa$, then decreased heating will force it outwards again. However, if the stagnation radius moves well outside of $\rIa$, then the high level of Ia heating will force it inwards.

Our criterion for cooling instability given by equations \eqref{eq:cooling2} and ~\eqref{eq:Mdotmax} is no longer valid if $\rs$ is regulated to $\rIa$ instead of being determined by equation \eqref{eq:stag_simple}.  However, by re-evaluating the density and temperature of gas at $\rIa$ using equations  (\ref{eq:rhors2}) and (\ref{eq:Tanalytic}) for $v_w = v_{w}^{\rm Ia}$, we the ratio of heating to radiative cooling at $\rIa$ is given by
\begin{align}
&\left.\frac{\dot{q}_{\rm heat}}{\left|\dot{q}_{\rm rad}\right|}\right|_{\rIa}\simeq
\begin{cases}
  90 \Mbheight^{1.3} (\epsilon_{\rm Ia}/0.4)^{1.7} (\tau_{\star}/\th)^{-3}
  \eta_{0.02}^{-2.7} & \text{core}\nonumber\\
   90 \Mbheight^{0.8} (\epsilon_{\rm Ia}/0.4)^{1.7} (\tau_{\star}/\th)^{-2.7}
  \eta_{0.02}^{-2.7} & \text{cusp}\nonumber\\
\end{cases}\\&\simeq
\begin{cases}
  90 \Mbheight^{1.3} (\epsilon_{\rm Ia}/0.4)^{1.7}
  (\tau_{\star}/\th)^{0.5}
 & \text{core}\\
   90 \Mbheight^{0.8} (\epsilon_{\rm Ia}/0.4)^{1.7} (\tau_{\star}/\th)^{0.9}
  & \text{cusp}\\
\end{cases}
\end{align}
Thus, only for high mass galaxies and/or relatively old stellar populations is Ia heating is sufficient to prevent cooling instabilities on scales of the stagnation radius.
